Transformation of forecasts for evaluating predictive performance in an epidemiological context

### 2023-01-23 main.tex

\title{Transformation of forecasts for evaluating predictive performance in an epidemiological context}
  \author{Nikos I. Bosse\thanks{Department of Infectious Disease Epidemiology, London School of Hygiene \& Tropical Medicine, London, United Kingdom} $^{,}$\thanks{Centre for the Mathematical Modelling of Infectious Diseases, London, United Kingdom} $^{,*}$, 
  Sam Abbott\footnotemark[1] $^{,}$\footnotemark[2]$ ^{}$, 
  Anne Cori\thanks{MRC Centre for Outbreak Analysis and Modelling, Department of Infectious Disease Epidemiology, School of Public Health, Imperial College London, London, United Kingdom} $^{}$, \\
  Edwin van Leeuwen\thanks{Department of Infectious Disease Epidemiology, London School of Hygiene \& Tropical Medicine, London, United Kingdom} $^{,}$\thanks{Modelling & Economics Unit and NIHR Health Protection Research Unit in Modelling \& Health Economics, UK Health Security Agency, London, United Kingdom} $^{}$, 
  Johannes Bracher\thanks{Chair of Statistical Methods and Econometrics, Karlsruhe Institute of Technology, Karlsruhe, Germany } $^{,}$\thanks{Computational Statistics Group, Heidelberg Institute for Theoretical Studies, Heidelberg, Germany } $^{, \dagger}$, 
  Sebastian Funk\footnotemark[1] $^{, }$\footnotemark[2]$ ^{}$ $ ^{, \dagger}$}

\maketitle
%\tableofcontents
\begin{abstract}

%I swapped the commented out version and this one on Wednesday, January 18 2.47pm EST. 
%Forecast evaluation plays an essential role in improving forecasting models and can support their role in informing decision-making. In epidemiology, probabilistic forecasts are often scored using the Weighted Interval Score (WIS) or the Continuous Ranked Probability Score (CRPS), which are measures of absolute distance between forecast and observation. Infectious disease transmission is commonly modelled as an exponential process characterized by the growth rate or the effective reproduction number. However, if the true objective of the forecaster is to predict future growth rates or reproduction numbers, then scoring based on the absolute distance between forecast and observation may not reflect this underlying task well and thus make it difficult to compare forecasts across time, space, and forecast targets. As a solution, we propose transforming the data before scoring the forecasts to better reflect the underlying generative process. Specifically, we discuss applying CRPS after log transformation of predictions and observations, which can be interpreted as a measure of relative error and as a score for an implicit forecast of the exponential growth rate. Moreover, we discuss its possible role as a variance-stabilizing transformation. Applying the natural logarithm to data from the European COVID-19 Forecast Hub, we found that it changed model rankings, regardless of stratification by forecast date, location or target types, and emphasised situations in which models missed lower inflection points and consequent rises in numbers, while penalising missed peaks less severely. We conclude that appropriate transformations, of which the natural logarithm is only one possible option, should be considered when assessing the performance of different models in the context of infectious disease incidence.

@@

For conceptual clarity and to allow for a more in-depth discussion, we focus mostly on the natural logarithm as a particular transformation (referred to as the log-transformtation in the remainder of this manuscript) in the context of epidemic phenomena. Instead of a score representing the magnitude of absolute errors, applying a log-transformation prior to the CRPS yields a score which a) measures relative error (see Section \ref{sec:methods:relative}), b) provides a measure for how well a forecast captures the exponential growth rate of the target quantity (see Section \ref{sec:methods:growthrate}) and c) is less dependent on the expected order of magnitude of the quantity to be predicted (see Section \ref{sec:methods:vst}). %The interpretation of the log-transformed score as a score for a forecast of the exponential growth rate of a given quantity aligns the evaluation with our understanding of how epidemiological processes should be modeled and the kinds of errors models would typically make. 
We therefore argue that such evaluations on the logarithmic scale should complement the prevailing evaluations on the natural scale. 
Other transformations may likewise be of interest. We briefly explore the square root transformation as an alternative transformation. %[and population standardization?]. 
Our analysis mostly focuses on the CRPS (or WIS) as an evaluation metric for probabilistic forecasts, given its widespread use throughout the COVID-19 pandemic.

The remainder of the article is structured as follows. In Sections \ref{sec:methods:relative}--\ref{sec:methods:vst} we provide some mathematical intuition on applying the log-transformation prior to evaluating the CRPS, highlighting the connections to relative error measures, the epidemic growth rate and variance stabilizing transformations.
%We first provide a mathematical intuition of the impact of the log-transformation and its interpretation as relative error (Section \ref{sec:methods:relative}, measure of the accuracy of a forecast of the growth rate (Section \ref{sec:methods:growthrate}) and it's effect as a variance-stabilising transformation that makes can make scores independent of the mean of the forecast target (Section \ref{sec:methods:vst}. 
We then discuss practical considerations for applying transformations in general and the log-transformation in particular (Section \ref{sec:methods:considerations}) and the effect of the log-transformation on forecast rankings (Section \ref{sec:methods:rankings}). To analyse the real-world implications of the log-transformation we use forecasts submitted to the European COVID-19 Forecast Hub  \citep[Section \ref{sec:HUB}]{europeancovid-19forecasthubEuropeanCovid19Forecast2021, sherrattPredictivePerformanceMultimodel2022}. Finally, we provide scoring recommendations, discuss alternative transformations that may be useful in different contexts, and suggest further research avenues (Section \ref{sec:discussion}).

@@

%The analogy between the WIS on the log scale and the absolute error of a point forecast on the log scale is slightly inaccurate: while the absolute error on the log scale is exactly symmetric (e.g., predicting 2 instead of 1 would give the same relative error as predicting 0.5), this is not entirely true for the WIS, which is computed for a set of prediction intervals. Due to the fact that upper and lower bounds of individual prediction intervals are affected slightly differently by the log-transformations, small deviations may occur (compare e.g. in Figure \ref{fig:change-in-scores}D the small difference between scores for an observed outcome of $\frac{1}{5}$ and 5).

@@

When evaluating models across sets of forecasting tasks, it may be desirable for each target to have a similar impact on the overall results. In disease incidence forecasting, this is not the case when using the CRPS on the natural scale, as the latter typically scales with the order of magnitude of the quantity to be predicted. Average scores are then dominated by the results achieved for targets with high expected outcomes. 
%Bracher et al. mention this as a desirable feature of the CRPS. 
% In epidemiological contexts, the variance of the variable of interest usually scales with its mean. 
Specifically, if the predictive distribution for the quantity $Y$ equals the true data-generating process $F$ (an ideal forecast), the expected CRPS is given by \citep{gneitingStrictlyProperScoring2007}
$$
\mathbb{E}[\text{CRPS}(F, y)] = 0.5\times\mathbb{E}|Y - Y'|,
$$
where $Y$ and $Y'$ are independent samples from $F$. This corresponds to half the \textit{mean absolute difference}, which is a measure of dispersion. If $F$ is well-approximated by a normal distribution $\text{N}(\mu, \sigma^2)$, the approximation
$$
\mathbb{E}_F[\text{CRPS}(F, y)] \approx \frac{\sigma}{\sqrt{\pi}}
$$
can be used. This means that the expected CRPS scales roughly with the standard deviation, which in turn typically increases with the mean in epidemiological forecasting. In order to make the expected CRPS independent of the expected outcome, a \textit{variance-stabilising transformation} \citep[VST,][]{bartlettSquareRootTransformation1936} can be employed. The choice of this transformation depends on the mean-variance relationship of the underlying process.

@@

In this paper, we proposed the use of transformations, with a particular focus on the natural logarithmic transformation, when evaluating forecasts in an epidemiological setting. These transformations can address issues that arise when evaluating epidemiological forecasts based on measures of absolute error and their probabilistic generalisations (i.e CRPS and WIS). We showed that scores obtained after log-transforming both forecasts and observations can be interpreted as a) a measure of relative prediction errors, as well as b) a score for a forecast of the exponential growth rate of the target quantity and c) as variance stabilising transformation in some settings.
When applying this approach to forecasts from the European COVID-19 Forecast Hub, we found overall scores on the log scale to be more equal across, time, location and target type (cases, deaths) than scores on the natural scale. Scores on the log scale were much less influenced by the overall incidence level in a country and showed a slight tendency to be higher in locations with very low incidences. We found that model rankings changed noticeably. 
%, favouring models like epiforecasts-EpiNow2 which uses a log scale generative process and a negative binomial observation model, and disfavouring models like MUNI-ARIMA which uses a square root transform to stabilise variance. 
On the natural scale, missing the peak and overshooting was more severely penalised than missing the nadir and the following upswing in numbers. Both failure modes tended to be more equally penalised on the log scale (with undershooting receiving slightly higher penalties in our example).

>>>>

\title{Transformation of forecasts for evaluating predictive performance in an epidemiological context}
  \author{Nikos I. Bosse\thanks{Department of Infectious Disease Epidemiology, London School of Hygiene \& Tropical Medicine, London, United Kingdom} $^{,}$\thanks{Centre for the Mathematical Modelling of Infectious Diseases, London, United Kingdom} $^{,}$\thanks{NIHR Health Protection Research Unit in Modelling \& Health Economics} $^{,*}$,  
  Sam Abbott\footnotemark[1] $^{,}$\footnotemark[2]$ ^{}$, 
  Anne Cori\thanks{MRC Centre for Outbreak Analysis and Modelling, Department of Infectious Disease Epidemiology, School of Public Health, Imperial College London, London, United Kingdom} $^{}$, \\
  Edwin van Leeuwen\footnotemark[1] $^{,}$\thanks{Modelling \& Economics Unit and NIHR Health Protection Research Unit in Modelling \& Health Economics, UK Health Security Agency, London, United Kingdom} $^{}$, 
  Johannes Bracher\thanks{Chair of Statistical Methods and Econometrics, Karlsruhe Institute of Technology, Karlsruhe, Germany } $^{,}$\thanks{Computational Statistics Group, Heidelberg Institute for Theoretical Studies, Heidelberg, Germany } $^{, \dagger}$, 
  Sebastian Funk\footnotemark[1] $^{, }$\footnotemark[2] $^{, }$\footnotemark[3] $ ^{, \dagger}$}

\maketitle
\begin{abstract}

@@

For conceptual clarity and to allow for a more in-depth discussion, we focus mostly on the natural logarithm as a particular transformation (referred to as the log-transformtation in the remainder of this manuscript) in the context of epidemic phenomena. Instead of a score representing the magnitude of absolute errors, applying a log-transformation prior to the CRPS yields a score which a) measures relative error (see Section \ref{sec:methods:relative}), b) provides a measure for how well a forecast captures the exponential growth rate of the target quantity (see Section \ref{sec:methods:growthrate}) and c) is less dependent on the expected order of magnitude of the quantity to be predicted (see Section \ref{sec:methods:vst}). 
We therefore argue that such evaluations on the logarithmic scale should complement the prevailing evaluations on the natural scale. 
Other transformations may likewise be of interest. We briefly explore the square root transformation as an alternative transformation. 
Our analysis mostly focuses on the CRPS (or WIS) as an evaluation metric for probabilistic forecasts, given its widespread use throughout the COVID-19 pandemic.

The remainder of the article is structured as follows. In Sections \ref{sec:methods:relative}--\ref{sec:methods:vst} we provide some mathematical intuition on applying the log-transformation prior to evaluating the CRPS, highlighting the connections to relative error measures, the epidemic growth rate and variance stabilizing transformations.
We then discuss practical considerations for applying transformations in general and the log-transformation in particular (Section \ref{sec:methods:considerations}) and the effect of the log-transformation on forecast rankings (Section \ref{sec:methods:rankings}). To analyse the real-world implications of the log-transformation we use forecasts submitted to the European COVID-19 Forecast Hub  \citep[ Section \ref{sec:HUB}]{europeancovid-19forecasthubEuropeanCovid19Forecast2021, sherrattPredictivePerformanceMultimodel2022}. Finally, we provide scoring recommendations, discuss alternative transformations that may be useful in different contexts, and suggest further research avenues (Section \ref{sec:discussion}).

@@

@@

When evaluating models across sets of forecasting tasks, it may be desirable for each target to have a similar impact on the overall results. In disease incidence forecasting, this is not the case when using the CRPS on the natural scale, as the latter typically scales with the order of magnitude of the quantity to be predicted. Average scores are then dominated by the results achieved for targets with high expected outcomes.

Specifically, if the predictive distribution for the quantity $Y$ equals the true data-generating process $F$ (an ideal forecast), the expected CRPS is given by \citep{gneitingStrictlyProperScoring2007}
$$
\mathbb{E}[\text{CRPS}(F, y)] = 0.5\times\mathbb{E}|Y - Y'|,
$$
where $Y$ and $Y'$ are independent samples from $F$. This corresponds to half the \textit{mean absolute difference}, which is a measure of dispersion. If $F$ is well-approximated by a normal distribution $\text{N}(\mu, \sigma^2)$, the approximation
$$
\mathbb{E}_F[\text{CRPS}(F, y)] \approx \frac{\sigma}{\sqrt{\pi}}
$$
can be used. This means that the expected CRPS scales roughly with the standard deviation, which in turn typically increases with the mean in epidemiological forecasting. In order to make the expected CRPS independent of the expected outcome, a \textit{variance-stabilising transformation} \citep[VST,][]{bartlettSquareRootTransformation1936} can be employed. The choice of this transformation depends on the mean-variance relationship of the underlying process.

@@

In this paper, we proposed the use of transformations, with a particular focus on the natural logarithmic transformation, when evaluating forecasts in an epidemiological setting. These transformations can address issues that arise when evaluating epidemiological forecasts based on measures of absolute error and their probabilistic generalisations (i.e CRPS and WIS). We showed that scores obtained after log-transforming both forecasts and observations can be interpreted as a) a measure of relative prediction errors, as well as b) a score for a forecast of the exponential growth rate of the target quantity and c) as variance stabilising transformation in some settings.
When applying this approach to forecasts from the European COVID-19 Forecast Hub, we found overall scores on the log scale to be more equal across, time, location and target type (cases, deaths) than scores on the natural scale. Scores on the log scale were much less influenced by the overall incidence level in a country and showed a slight tendency to be higher in locations with very low incidences. We found that model rankings changed noticeably.

On the natural scale, missing the peak and overshooting was more severely penalised than missing the nadir and the following upswing in numbers. Both failure modes tended to be more equally penalised on the log scale (with undershooting receiving slightly higher penalties in our example).
